---
title: "Building, checking, and diagramming a theory"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Building, checking, and diagramming a theory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`theoryforge` treats a scientific theory as a versioned, machine-checkable
object. This vignette walks through the core loop of building a theory,
checking its rigour and diagramming it. It is the R twin of the Python
package, and both follow the same pinned contract
([`API_SPEC.md`](https://github.com/pablobernabeu/theoryforge/blob/main/API_SPEC.md)).

```{r setup}
library(theoryforge)
```

## Building a theory

A theory is built incrementally with the BUILDING-mode verbs. Each verb appends
to the theory and records a provenance entry, so the steps compose cleanly with
the native pipe.

```{r build}
theory <- tf_theory("panic-network", "A network theory of panic") |>
  tf_add_construct("c_arousal", "Physiological arousal",
                   "Bodily activation in response to a stressor.",
                   measurement = "heart rate variability",
                   boundary_conditions = "awake adults") |>
  tf_add_construct("c_threat", "Perceived threat",
                   "Appraised danger of bodily sensations.",
                   measurement = "self-report appraisal scale",
                   boundary_conditions = "awake adults") |>
  tf_add_proposition("p1", "c_arousal", "c_threat", "causes",
                     mechanism = "Activation raises the salience of threat cues.") |>
  tf_add_prediction("h1", "Arousal raises threat appraisal by a fixed amount.",
                    "point", derives_from = "p1")

tf_validate(theory) # structural checks: required fields and enums
```

`tf_validate()` returns `TRUE` invisibly on success and stops with a message
listing every problem otherwise. It mirrors the Python `Theory.validate` and
does not require any optional dependency.

A theory can be written to and read back from disk. The format follows the file
extension (`.json` for JSON, otherwise YAML).

```{r io}
path <- tempfile(fileext = ".yaml")
tf_write(theory, path)
roundtrip <- tf_read(path)
identical(roundtrip$id, theory$id)
```

## Checking rigour

`tf_check()` runs the 12-item rigour checklist defined in the vendored
`rigor_checklist.yaml` and scored per
[`API_SPEC.md`](https://github.com/pablobernabeu/theoryforge/blob/main/API_SPEC.md) section 4.

```{r check}
report <- tf_check(theory)

report$aggregate_score   # weighted score, 0-100, rounded to 1 dp
report$gate              # "pass", "blocked", or "advisory" (draft maturity)
report$n_blockers_failed # count of failed blocker items

# Per-item detail (checklist order preserved):
report$items[[1]]$id      # "falsifiability"
report$items[[1]]$status  # "pass" / "warn" / "fail"
report$items[[1]]$score   # numeric in [0, 1]
```

Render it as JSON (valid, pretty-printed) or as an HTML fragment:

```{r report}
cat(tf_report(theory, format = "json"))
```

### Screening for redundant constructs

The lexical redundancy screen reports the Jaccard similarity of every pair of
construct definitions, sorted by descending similarity.

```{r redundancy}
tf_redundancy_check(theory)
```

## Diagramming

`tf_diagram()` emits byte-identical intermediate representations for several
diagram types. The digraphs are Graphviz DOT, and the causal DAG uses dagitty
syntax.

```{r diagram}
cat(tf_diagram(theory, type = "nomological_net"))
cat(tf_diagram(theory, type = "causal_dag"))
```

These strings are compared byte-for-byte against the Python reference outputs
in the package's tests, guaranteeing cross-language parity.

## Simulation

`tf_simulate()` treats each construct as a state variable and integrates the
signed proposition network as a linear dynamical system. The trajectory is
deterministic and parity-tested against the Python reference:

```{r simulate}
sim <- tf_simulate(theory, steps = 5)
sim$states
sim$trajectory[[1]]
```
