---
title: "Developing and testing a theory"
output: rmarkdown::html_vignette
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(theoryforge)
```

The companion vignette covers building a theory and checking its rigor. This
article continues the workflow into development and testing. It shows how to
operationalise severity, export a preregistration, appraise an amendment in
Lakatosian terms, compile a measurement model for `lavaan`, and assemble a
reviewer-facing audit bundle. The worked example reads the panic-network
fixture shipped with the package, so every chunk runs with only `theoryforge`
loaded.

```{r read}
theory <- tf_read(
  system.file("fixtures/panic-network.theory.yaml", package = "theoryforge")
)
theory$id
```

## Operationalised severity

A prediction earns evidential weight in proportion to the risk it takes. A
point prediction stakes more on the data than a directional one, and that
asymmetry should be made explicit before any test is run. `tf_severity()`
reports, for each prediction in file order, the riskiness of its claim form and
the resulting computed severity.

```{r severity}
tf_severity(theory)
```

The `risk_score` reflects the prediction type, while `computed_severity`
applies the discounts and bonuses defined in the specification. Reading these
values before data collection guards against the temptation to read a weak,
hard-to-fail prediction as though it were a stringent test.

## Preregistration

`tf_preregister()` renders a preregistration document as a single markdown
string. It records the constructs, propositions, and predictions in a fixed
order, so the output is stable across runs and suitable for deposit.

```{r prereg}
prereg <- tf_preregister(theory)
cat(prereg)
```

Passing a `path` writes the same markdown to disk with LF line endings, which
is convenient when the document is committed to a repository or uploaded to a
registry.

```{r prereg-write}
prereg_path <- tempfile(fileext = ".md")
invisible(tf_preregister(theory, path = prereg_path))
file.exists(prereg_path)
```

## Appraising an amendment

Theories change. The question is whether a change is progressive, in that it
predicts and then corroborates something new, or degenerating, in that it only
adds assumptions to absorb anomalies. `tf_appraise_amendment()` compares an
amended theory against its prior version and returns a verdict in those terms.

Here the prior is the panic-network fixture and the amended version is its 2026
revision, which adds one new prediction that is later corroborated and
introduces no new ad-hoc assumptions.

```{r appraise}
v2 <- tf_read(
  system.file("fixtures/panic-network-2026-v2.theory.yaml", package = "theoryforge")
)

appraisal <- tf_appraise_amendment(v2, theory)
appraisal$verdict
appraisal$new_predictions
appraisal$corroborated_new
appraisal$ad_hoc_assumptions
```

A `progressive` verdict reflects new content that survived a test. Had the
amendment instead added assumptions tied to specific anomalies without
generating corroborated predictions, the verdict would record that as
degenerating.

## Compiling a measurement model

When constructs carry measurement indicators, `tf_compile_sem()` compiles them
into `lavaan` model syntax. Constructs with indicators become latent
measurement models with `=~`, and propositions become structural paths with
`~`, covariances with `~~`, or annotated moderation lines.

```{r sem}
cat(tf_compile_sem(theory))
```

The function returns a plain string and does not require `lavaan` to be
installed. The output can be passed directly to `lavaan::sem()` once the data
are available.

## The reviewer audit bundle

`tf_dossier()` assembles a single markdown document for a reviewer or editor.
It collects the header, the rigor-checklist table, the severity list, the
provenance trail, and the preregistration into one bundle, so a reader can
follow the theory from its claims through to its planned tests in one place.

```{r dossier}
dossier <- tf_dossier(theory)
cat(substr(dossier, 1, 600))
```

The full document is deterministic and byte-identical to the Python
implementation, so the same theory produces the same dossier regardless of
which package generated it.
