---
title: "Developing and testing a theory"
output: rmarkdown::html_vignette
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(theoryforge)
```

The companion vignette covers building a theory and checking its rigour. This
article continues the workflow into development and testing. It shows how to
operationalise severity, export a preregistration, appraise an amendment in
Lakatosian terms, compile a measurement model for `lavaan`, and assemble a
reviewer-facing audit bundle. The worked example reads the panic-network
fixture shipped with the package, so every chunk runs with only `theoryforge`
loaded.

```{r read}
theory <- tf_read(
  system.file("fixtures/panic-network.theory.yaml", package = "theoryforge")
)
theory$id
```

## Operationalised severity

A prediction earns evidential weight in proportion to the risk it takes. A
point prediction stakes more on the data than a directional one, and that
asymmetry should be made explicit before any test is run. `tf_severity()`
reports, for each prediction in file order, the riskiness of its claim form and
the resulting computed severity.

```{r severity}
tf_severity(theory)
```

The `risk_score` reflects the prediction type, while `computed_severity`
applies the discounts and bonuses defined in the specification. Reading these
values before data collection guards against the temptation to read a weak,
hard-to-fail prediction as though it were a stringent test.

## Preregistration

`tf_preregister()` renders a preregistration document as a single markdown
string. It records the predictions, each with its claim type and derivation,
plus the severity values, in a fixed order, so the output is stable across
runs and suitable for deposit. Because the string is markdown, it renders
below as the document a registry would receive.

```{r prereg, results = "asis"}
prereg <- tf_preregister(theory)
cat(prereg)
```

Passing a `path` writes the same markdown to disk with LF line endings, which
is convenient when the document is committed to a repository or uploaded to a
registry.

```{r prereg-write}
prereg_path <- tempfile(fileext = ".md")
invisible(tf_preregister(theory, path = prereg_path))
file.exists(prereg_path)
```

## Appraising an amendment

Theories change. The question is whether a change is progressive, in that it
predicts and then corroborates something new, or degenerating, in that it only
adds assumptions to absorb anomalies. `tf_appraise_amendment()` compares an
amended theory against its prior version and returns a verdict in those terms.

Here the prior is the panic-network fixture and the amended version is its 2026
revision, which adds one new prediction that is later corroborated and
introduces no new ad-hoc assumptions.

```{r appraise}
v2 <- tf_read(
  system.file("fixtures/panic-network-2026-v2.theory.yaml", package = "theoryforge")
)

appraisal <- tf_appraise_amendment(v2, theory)
appraisal$verdict
appraisal$new_predictions
appraisal$corroborated_new
appraisal$ad_hoc_assumptions
```

A `progressive` verdict reflects new content that survived a test. Had the
amendment instead added assumptions tied to specific anomalies without
generating corroborated predictions, the verdict would record that as
degenerating.

## Compiling a measurement model

When constructs carry measurement indicators, `tf_compile_sem()` compiles them
into `lavaan` model syntax. Constructs with indicators become latent
measurement models with `=~`, and propositions become structural paths with
`~`, covariances with `~~` or annotated moderation lines.

```{r sem, collapse = FALSE, comment = ""}
cat(tf_compile_sem(theory))
```

The function returns a plain string and does not require `lavaan` to be
installed. The output can be passed directly to `lavaan::sem()` once the data
are available.

## The reviewer audit bundle

`tf_dossier()` assembles a single markdown document for a reviewer or editor.
It collects the header, the rigour-checklist table, the severity list, the
provenance trail and the preregistration into one bundle, so a reader can
follow the theory from its claims through to its planned tests in one place.
The opening of the bundle, through the rigour-checklist table, renders below.

```{r dossier, results = "asis"}
dossier <- tf_dossier(theory)
cat(strsplit(dossier, "\n## Severity", fixed = TRUE)[[1]][1])
```

The severity list, the provenance trail and the preregistration follow in the
full document, which is deterministic, so the same theory always produces the
same dossier.

## Rendering and depositing

`tf_render_report()` writes the dossier as a Quarto source document, ready to
render to HTML or PDF. With `render = FALSE` it writes the `.qmd` without calling
Quarto, so the step runs anywhere.

```{r render-report}
report_path <- tf_render_report(theory, tempfile(fileext = ".qmd"))
basename(report_path)
```

`tf_osf_push()` deposits the rendered report to an Open Science Framework node.
It makes an authenticated network call, so it sits outside the deterministic
core and is shown here without being run.

```{r osf, eval = FALSE}
tf_osf_push(theory, node = "ab12c", token = Sys.getenv("OSF_PAT"))
```

## Visualising the theory

`tf_diagram()` exports several views of the same object. The `workflow` view
traces the lifecycle from constructs, through propositions and predictions, to
the recorded test outcomes, so the path from a claim to its planned test is
visible in one figure.

```{r workflow, collapse = FALSE, comment = ""}
cat(tf_diagram(theory, "workflow"))
```

Rendered with `tf_render_diagram()`, which passes the DOT to the DiagrammeR
engine, the same view reads as a figure.

```{r workflow-render, results = "asis", eval = requireNamespace("DiagrammeR", quietly = TRUE) && requireNamespace("DiagrammeRsvg", quietly = TRUE)}
cat('<div class="tf-figure tf-diagram">', tf_render_diagram(theory, "workflow", as = "svg"), '</div>', sep = "")
```

The `context` view places the theory among the scope conditions under which it
is claimed to hold and the registered rivals it is meant to outpredict.

```{r context, collapse = FALSE, comment = ""}
cat(tf_diagram(theory, "context"))
```

```{r context-render, results = "asis", eval = requireNamespace("DiagrammeR", quietly = TRUE) && requireNamespace("DiagrammeRsvg", quietly = TRUE)}
cat('<div class="tf-figure tf-diagram">', tf_render_diagram(theory, "context", as = "svg"), '</div>', sep = "")
```

The `pipeline` view links each prediction to its recorded test outcome, so
predictions still awaiting a test are visible as loose ends.

```{r pipeline, collapse = FALSE, comment = ""}
cat(tf_diagram(theory, "pipeline"))
```

```{r pipeline-render, results = "asis", eval = requireNamespace("DiagrammeR", quietly = TRUE) && requireNamespace("DiagrammeRsvg", quietly = TRUE)}
cat('<div class="tf-figure tf-diagram">', tf_render_diagram(theory, "pipeline", as = "svg"), '</div>', sep = "")
```

The `provenance` view draws the build log as a digraph, so the record of how
the theory reached its current state travels with the object itself.

```{r provenance, collapse = FALSE, comment = ""}
cat(tf_diagram(theory, "provenance"))
```

```{r provenance-render, results = "asis", eval = requireNamespace("DiagrammeR", quietly = TRUE) && requireNamespace("DiagrammeRsvg", quietly = TRUE)}
cat('<div class="tf-figure tf-diagram">', tf_render_diagram(theory, "provenance", as = "svg"), '</div>', sep = "")
```

The `venn` view returns an SVG that shows where the constructs share boundary
conditions. It renders inline because the output is a self-contained image.

```{r venn, results = "asis"}
cat('<div class="tf-figure">', tf_diagram(theory, "venn"), '</div>', sep = "")
```

Two further views summarise the scoring. The `rigour` view draws the checklist
as a status grid, colouring each item by its result and reporting the aggregate
score and the gate.

```{r rigour-view, results = "asis"}
cat('<div class="tf-figure">', tf_diagram(theory, "rigour"), '</div>', sep = "")
```

The `severity` view draws one bar per prediction, scaled by its computed
severity, so the riskier tests stand out at a glance.

```{r severity-view, results = "asis"}
cat('<div class="tf-figure">', tf_diagram(theory, "severity"), '</div>', sep = "")
```

The `development_roadmap` view turns the same checklist into a worklist by
keeping only the items that still fail or warn. The panic network passes every
check, so its roadmap collapses to a single node, and the deliberately weak
theory shipped alongside it shows the worklist in full.

```{r roadmap, collapse = FALSE, comment = ""}
weak <- tf_read(
  system.file("fixtures/weak-theory.theory.yaml", package = "theoryforge")
)
cat(tf_diagram(theory, "development_roadmap"))
cat(tf_diagram(weak, "development_roadmap"))
```

```{r roadmap-render, results = "asis", eval = requireNamespace("DiagrammeR", quietly = TRUE) && requireNamespace("DiagrammeRsvg", quietly = TRUE)}
cat('<div class="tf-figure tf-diagram">', tf_render_diagram(weak, "development_roadmap", as = "svg"), '</div>', sep = "")
```

All of these views are deterministic. The DOT strings render with any Graphviz
engine, and the SVG views open in any browser. Together with the
`nomological_net` and `causal_dag` views introduced in the Get started
vignette, they complete the set of diagram types that `tf_diagram()` exports,
each documented in `?tf_diagram`.
