---
title: "Developing and testing a theory"
output: rmarkdown::html_vignette
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "")
# Console colour carries no meaning on a rendered page. pkgdown turns it on for
# its own build, and the escape sequences then reach the reader as literal text,
# so colour is switched off here for a plain vignette render and a site build
# alike. The fixed width keeps printed output inside the documentation column.
options(cli.num_colors = 1, cli.hyperlink = FALSE, crayon.enabled = FALSE,
        width = 80)
# Both optional packages are needed to turn the DOT source into an inline
# SVG, so the rendering chunks below are skipped when either is absent.
has_diagram <- requireNamespace("DiagrammeR", quietly = TRUE) &&
  requireNamespace("DiagrammeRsvg", quietly = TRUE)
library(theoryforge)
```

The companion vignette covers building a theory and checking its rigour. This
article continues the workflow into development and testing. It shows how to
grade the riskiness of each prediction's claim, derive what the causal graph
forbids in data, export a preregistration, appraise an amendment in Lakatosian
terms, compile a measurement model for `lavaan` and assemble a reviewer-facing
audit bundle. The worked example reads the panic-network fixture shipped with
the package, joined in one section by the shipped modality-switching fixture,
so every chunk runs with only `theoryforge` loaded.

```{r read}
theory <- tf_read(
  system.file("fixtures/panic-network.theory.yaml", package = "theoryforge")
)
theory$id
```

## The severity rubric

A prediction earns evidential weight in proportion to the risk it takes. A
point prediction stakes more on the data than a directional one, and that
asymmetry should be made explicit before any test is run. `tf_severity()` is a
pre-data rubric of claim form. It reports, for each prediction in file order,
the riskiness of its declared type and the resulting computed severity.

```{r severity}
tf_severity(theory)
```

The `risk_score` reflects the prediction type, while `computed_severity`
applies the directional discount and the diagnostic bonus defined in the
specification, both of them the package's conventions. The rubric reads only
the declared type and the named rivals, never the statement or the data, so it
says nothing about how severely a claim has been tested. That depends on the
design and the data. Reading these values before data collection guards against
the temptation to read a weak, hard-to-fail prediction as though it were a risky
one.

## What the causal graph commits you to

Severity scores the predictions a theory states. A causal theory also commits
itself to claims it never states, because the propositions taken together are a
graph, and the graph implies a set of conditional independencies.
`tf_implications()` derives them. Each directed relation (increases, decreases,
causes, mediates and moderates) is an arrow, and an association is a
bidirected edge, covariance the theory leaves unexplained. Every pair of
constructs with no edge between them yields one claim when some set of other
constructs separates them, usually the parents of both. When every relation is
directed, those claims are the basis set, from which every other independence
the graph implies follows. Each claim is something a study can be designed to
look for and fail to find. An association can leave a pair that no set
separates, and `tf_implications()` lists such a pair under `inseparable`, since
the theory then implies no independence for it.

The panic-network example gets no statements. Its third proposition returns
from perceived threat to arousal, closing the feedback loop the theory is about,
and `tf_implications()` refuses a graph with a cycle.

```{r implications-cycle, error = TRUE}
tf_implications(theory)
```

The refusal names the cycle it found, so nobody has to hunt for it. A cyclic
graph does imply conditional independencies, and dagitty accepts the
`causal_dag` export of one. dagitty reads it by d-separation, though, which a
model with feedback is guaranteed to satisfy only in special cases, a linear
model among them. The criterion that holds whenever each feedback loop has a
unique equilibrium is sigma-separation
([Bongers et al., 2021](https://doi.org/10.1214/21-AOS2064)), and
`tf_implications()` does not apply it. `tf_simulate()` runs the loop over time.

The package ships a second example whose causal graph is acyclic, a theory of
modality switching in grounded conceptual processing. Sensorimotor experience
with a concept drives activation of the modality-specific perceptual system.
That activation both raises the cost of switching modality between consecutive
trials and eases conceptual access, which lexical familiarity with the word form
eases as well.

```{r implications}
switching <- tf_read(
  system.file("fixtures/modality-switching.theory.yaml", package = "theoryforge")
)

implied <- tf_implications(switching)
implied$n_edges
implied$n_implications
writeLines(vapply(implied$implications, function(i) i$statement, character(1)))
```

Five constructs and four causal propositions leave six non-adjacent pairs, and
so six claims, each written in the notation dagitty prints. Two of them are
worth dwelling on. Switch cost and ease of conceptual access are the two
children of a fork at modality activation, so the theory says the correlation
between them is entirely due to that shared cause and vanishes once it is held
fixed. Sensorimotor experience and lexical familiarity meet only at the collider at
conceptual access, and the theory says they are unrelated to each other with
nothing held fixed, which is what an account deriving perceptual knowledge from
distributional word statistics would deny. Neither
claim appears among the theory's own predictions, and either could sink it. A
version of the theory that granted sensorimotor experience and lexical
familiarity some covariance would state it with `associates`, and the claim that
the two are unrelated would then be withdrawn.

The claims concern constructs, and a study observes them through fallible
measures. Error in a conditioning construct leaves part of the dependence that
holding it fixed should remove. A conditional claim tested on observed scores is
therefore rejected too often, and more often the larger the sample
([Westfall & Yarkoni, 2016](https://doi.org/10.1371/journal.pone.0152719)). Take
a chain whose two paths have standardised coefficients of .5, with the middle
construct measured at a reliability of .8. A partial-correlation test at the 5
per cent level then rejects the true claim in about 14, 29 and 51 per cent of
studies of 200, 500 and 1,000 observations. Five of the six claims above are
conditional. They are better tested in a latent-variable model, such as one
built on the measurement model that `tf_compile_sem()` writes
([Thoemmes et al., 2018](https://doi.org/10.1037/met0000147)).

## Preregistration

`tf_preregister()` renders a preregistration document as a single markdown
string. It records the predictions, each with its claim type and derivation,
plus the severity values, in a fixed order, so the output is stable across
runs and suitable for deposit. Because the string is markdown, it renders
below as the document a registry would receive.

```{r prereg, results = "asis"}
prereg <- tf_preregister(theory)
cat(prereg)
```

Passing a `path` writes the same markdown to disk with LF line endings, which
is convenient when the document is committed to a repository or uploaded to a
registry.

```{r prereg-write}
prereg_path <- tempfile(fileext = ".md")
invisible(tf_preregister(theory, path = prereg_path))
file.exists(prereg_path)
```

## Appraising an amendment

Theories change. The question is whether a change is progressive, in that it
predicts and then corroborates something new, or degenerating, in that it only
adds assumptions to absorb anomalies. `tf_appraise_amendment()` compares an
amended theory against its prior version and returns a verdict in those terms.

Here the prior is the panic-network fixture and the amended version is its 2026
revision. The revision adds a proposition, a direct path from arousal to
avoidance, and derives from it a new prediction that a registered test later
corroborates. It introduces no ad hoc assumption.

```{r appraise}
v2 <- tf_read(
  system.file(
    "fixtures/panic-network-2026-v2.theory.yaml",
    package = "theoryforge"
  )
)

appraisal <- tf_appraise_amendment(v2, theory)
appraisal$verdict
appraisal$new_predictions
appraisal$corroborated_new
appraisal$articulated
appraisal$ad_hoc_assumptions
```

A `progressive` verdict reflects new content that survived a test. The
appraisal compares what the two versions claim. A prediction that only changed
its identifier is reported under `renamed`, and one derived only from
propositions the prior already held is reported under `articulated` and does
not make the amendment progressive. Had the amendment instead added
assumptions tied to specific anomalies without generating corroborated
predictions, the verdict would record that as degenerating.

## Compiling a measurement model

When constructs carry measurement indicators, `tf_compile_sem()` compiles them
into `lavaan` model syntax. Constructs with indicators become latent
measurement models with `=~`, and propositions become structural paths with
`~`, covariances with `~~` or annotated moderation lines.

```{r sem}
cat(tf_compile_sem(theory))
```

The function returns a plain string and does not require `lavaan` to be
installed. The output can be passed directly to `lavaan::sem()` once the data
are available.

## The reviewer audit bundle

`tf_dossier()` assembles a single markdown document for a reviewer or editor.
It collects the header, the rigour-checklist table, the severity list, the
provenance trail and the preregistration into one bundle, so a reader can
follow the theory from its claims through to its planned tests in one place.
The opening of the bundle, through the rigour-checklist table, renders below.

```{r dossier, results = "asis"}
dossier <- tf_dossier(theory)
cat(strsplit(dossier, "\n## Severity", fixed = TRUE)[[1]][1])
```

The severity list, the provenance trail and the preregistration follow in the
full document, which is deterministic, so the same theory always produces the
same dossier.

## Rendering and depositing

`tf_render_report()` writes the dossier as a Quarto source document, ready to
render to HTML or PDF. With `render = FALSE`, which is the default, it writes the
`.qmd` without calling Quarto, so the step runs anywhere. It returns the path of
the file it wrote.

```{r render-report}
report_path <- tf_render_report(theory, tempfile(fileext = ".qmd"))
basename(report_path)
```

`tf_osf_push()` deposits the audit dossier on an Open Science Framework node, as
`<id>.dossier.md`. It defaults to `dry_run = TRUE`, which assembles the request
and returns it without sending anything, so the planned deposit can be inspected
offline and without a token. What goes up is the dossier, since that is the
machine-checkable bundle a reviewer needs. The report written just above stays on
your own machine.

```{r osf}
str(tf_osf_push(theory, node = "ab12c"))
```

The returned request records the method, the destination URL, the filename the
deposit would create and the size of the dossier in bytes, so the step can be
checked before anything leaves the machine. A live deposit needs
`dry_run = FALSE` together with a token and the node id, and that is the only
path that touches the network.

```{r osf-live, eval = FALSE}
tf_osf_push(
  theory,
  node = "ab12c",
  token = Sys.getenv("OSF_PAT"),
  dry_run = FALSE
)
```

OSF refuses to create a file whose name already exists in the project, so
depositing the same theory a second time stops with HTTP 409. Either pass a
filename that names the version, or add `overwrite = TRUE`, which lists the
project folder first and sends the dossier to the existing file as a new OSF
version. Any other refusal also stops with its HTTP status, so a returned value
always describes an upload that OSF accepted.

## Visualising the theory

`tf_diagram()` exports several views of the same object. The `workflow` view
traces the lifecycle from constructs, through propositions and predictions, to
the recorded test outcomes, so the path from a claim to its planned test is
visible in one figure.

```{r workflow}
cat(tf_diagram(theory, "workflow"))
```

Rendered with `tf_render_diagram()`, which passes the DOT to the DiagrammeR
engine, the same view reads as a figure.

```{r workflow-render, results = "asis", eval = has_diagram}
cat(
  '<div class="tf-figure tf-diagram">',
  tf_render_diagram(theory, "workflow", as = "svg"),
  '</div>',
  sep = ""
)
```

The `context` view places the theory among the scope conditions under which it
is claimed to hold and the registered rivals it is meant to outpredict.

```{r context}
cat(tf_diagram(theory, "context"))
```

```{r context-render, results = "asis", eval = has_diagram}
cat(
  '<div class="tf-figure tf-diagram">',
  tf_render_diagram(theory, "context", as = "svg"),
  '</div>',
  sep = ""
)
```

The `pipeline` view links each prediction to its recorded test outcome, so
predictions still awaiting a test are visible as loose ends.

```{r pipeline}
cat(tf_diagram(theory, "pipeline"))
```

```{r pipeline-render, results = "asis", eval = has_diagram}
cat(
  '<div class="tf-figure tf-diagram">',
  tf_render_diagram(theory, "pipeline", as = "svg"),
  '</div>',
  sep = ""
)
```

The `provenance` view draws the build log as a digraph, so the record of how
the theory reached its current state travels with the object itself.

```{r provenance}
cat(tf_diagram(theory, "provenance"))
```

```{r provenance-render, results = "asis", eval = has_diagram}
cat(
  '<div class="tf-figure tf-diagram">',
  tf_render_diagram(theory, "provenance", as = "svg"),
  '</div>',
  sep = ""
)
```

The `venn` view takes each construct's boundary conditions as a set and returns
an SVG showing where those scopes coincide and where they part company, with
each region labelled by the number of conditions in it. It renders inline
because the output is a self-contained image.

```{r venn, results = "asis"}
cat('<div class="tf-figure">', tf_diagram(theory, "venn"), '</div>', sep = "")
```

Two further views summarise the scoring. The `rigour` view draws the checklist
as a status grid, colouring each item by its result and reporting the aggregate
score and the gate.

```{r rigour-view, results = "asis"}
cat('<div class="tf-figure">', tf_diagram(theory, "rigour"), '</div>', sep = "")
```

The `severity` view draws one bar per prediction, scaled by its computed
severity, so the riskier claims stand out.

```{r severity-view, results = "asis"}
cat(
  '<div class="tf-figure">',
  tf_diagram(theory, "severity"),
  '</div>',
  sep = ""
)
```

The `development_roadmap` view turns the same checklist into a worklist by
keeping only the items that still fail or warn, and it orders that worklist the
way the work should be done. Whatever gates the theory comes first, heaviest
check first, and the advisory items follow. Each step carries its number, the
criterion it is measured against and whether missing it blocks the gate or is
merely advisory, so the figure amounts to a plan of work. The hub names the
theory and its current standing. The panic network passes every check, so its
roadmap reduces to that hub and an `all checks pass` node, and the deliberately
weak theory shipped alongside it shows the worklist in full.

```{r roadmap}
weak <- tf_read(
  system.file("fixtures/weak-theory.theory.yaml", package = "theoryforge")
)
cat(tf_diagram(theory, "development_roadmap"))
cat(tf_diagram(weak, "development_roadmap"))
```

```{r roadmap-render, results = "asis", eval = has_diagram}
cat(
  '<div class="tf-figure tf-diagram">',
  tf_render_diagram(weak, "development_roadmap", as = "svg"),
  '</div>',
  sep = ""
)
```

All of these views are deterministic. The DOT strings render with any Graphviz
engine, and the SVG views open in any browser. Together with the
`nomological_net` and `causal_dag` views introduced in the Get started
vignette, they complete the set of diagram types that `tf_diagram()` exports,
each documented in `?tf_diagram`.
