---
title: "Mapping the literature"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A theory is rarely built in isolation. It sits within a field that already has
crowded areas, sparse areas and recurring sets of references. The literature
layer of `theoryforge` gives a deterministic, machine-checkable account of that
field, so the placement of a theory within it can be argued from evidence
rather than from impression. This article reads a small corpus, computes a
bibliometric map, positions a theory against the map, and emits a diagram. The
analysis functions are deterministic and require no optional dependency.

```{r setup}
library(theoryforge)
```

## Reading a corpus

A corpus is a simple object of the form `{schema_version, id, records}`, where
each record carries an `id` and may carry `title`, `year`, `keywords` and
`references`. The package ships a small fixture corpus around panic and anxiety
research. The format is chosen by file extension, so the same reader handles
YAML and JSON.

```{r read-corpus}
corpus_path <- system.file("fixtures/panic-corpus.yaml", package = "theoryforge")
corpus <- tf_read_corpus(corpus_path)

corpus$id
length(corpus$records)   # number of works in the corpus
corpus$records[[1]]       # first record
```

## Building a literature map

`tf_litmap()` computes three things from the records. It counts keyword
co-occurrence within records, groups co-occurring keywords into thematic
components, and counts reference co-citation. An edge is kept only when its
co-occurrence count reaches `min_link` (default `2`), which suppresses incidental
pairings. Records are read in file order and every result is sorted, so the map
is fully deterministic.

```{r litmap}
lm <- tf_litmap(corpus, min_link = 2)

lm$n_records
unlist(lm$keywords)                 # the vocabulary of the corpus
lm$keyword_cooccurrence[[1]]        # first kept co-occurrence edge
```

The thematic components are the connected groups of co-occurring keywords. The
fixture is constructed so that the threshold-two graph separates cleanly into
four themes.

```{r themes}
for (theme in lm$themes) {
  cat(theme$id, ":", paste(unlist(theme$keywords), collapse = ", "), "\n")
}
```

Co-citation is computed the same way over the `references` field. Two cited
sources are linked when at least `min_link` records in the corpus cite them
together.

```{r cocitation}
lm$co_citation
```

## Positioning a theory against the field

`tf_landscape()` maps a theory's focal constructs and its registered
alternatives onto the themes from `tf_litmap()`. A theme is tagged
`"under_theorised"` when neither the focal theory nor any alternative touches it,
`"covered"` when exactly one does, and `"crowded"` when two or more do. Matching
is by shared tokens between the theme keywords and the text of the construct
labels, the theory title and the alternatives.

We build a small theory whose constructs speak to arousal and to appraisal, and
register one alternative that also speaks to appraisal. This leaves the arousal
theme covered, the appraisal theme crowded and the remaining themes untouched.

```{r theory}
theory <- tf_theory("panic-appraisal", "Arousal and appraisal in panic") |>
  tf_add_construct("c_arousal", "Physiological arousal",
                   "Bodily activation in response to a stressor.") |>
  tf_add_construct("c_appraisal", "Threat appraisal",
                   "Catastrophic appraisal of bodily sensations.") |>
  tf_add_alternative("alt_cog", "Cognitive appraisal account",
                     key_constructs = c("appraisal"))

landscape <- tf_landscape(theory, corpus, min_link = 2)
landscape$theory_id
```

Each theme in the result records which alternatives land on it, whether the
focal theory lands on it, and the resulting status.

```{r landscape-themes}
for (theme in landscape$themes) {
  cat(theme$id,
      "| status:", theme$status,
      "| focal:", theme$focal,
      "| alternatives:", paste(unlist(theme$alternatives), collapse = ", "),
      "\n")
}
```

Two summaries draw out the practical reading. The under-theorised fronts are
themes that the field around this theory has not yet engaged, and so are
candidates for new work. The redundancy risk lists crowded themes, where a new
contribution would have to distinguish itself from existing accounts.

```{r landscape-summary}
unlist(landscape$under_theorised_fronts)
unlist(landscape$redundancy_risk)
```

## Emitting a literature diagram

`tf_lit_diagram()` renders a deterministic Graphviz DOT string for the
literature layer. The keyword co-occurrence and co-citation maps render as
undirected graphs, and a landscape result renders as a directed theme map. The
output is a plain string, so it can be written to a `.dot` file or passed to a
Graphviz renderer.

```{r diagram-keywords, collapse = FALSE, comment = ""}
cat(tf_lit_diagram(lm, type = "keyword_cooccurrence"))
```

The theme landscape diagram is produced from the `tf_landscape()` result rather
than from the map, since it carries the focal theory and the alternatives.

```{r diagram-landscape, collapse = FALSE, comment = ""}
cat(tf_lit_diagram(landscape, type = "theme_landscape"))
```

## Building a corpus from OpenAlex

The corpus above was read from a file. A corpus can also be assembled from the
OpenAlex works API with `tf_fetch_corpus()`. This adapter is assistive. It
makes a live network call and its result depends on an external service that
changes over time, so it sits outside the deterministic core of the package.
The call is shown here but not run, and the supplied email enters the OpenAlex
polite pool.

```{r fetch, eval = FALSE}
fetched <- tf_fetch_corpus("panic disorder interoception",
                           per_page = 25,
                           mailto = "me@example.org")
fetched_map <- tf_litmap(fetched)
```

Once fetched, the corpus object has the same shape as a file-read corpus, so it
flows into `tf_litmap()`, `tf_landscape()` and `tf_lit_diagram()` without any
further change. The recommended pattern is to fetch once, write the result to
disk, and then work from the saved file so that later analyses remain
reproducible.

## Tracking new evidence with an external search

Locating a corpus is only one use of a literature search. A second, recurring
need is narrower: checking whether a search has turned up any source the
theory does not already cite. `tf_new_evidence_dois()` answers that question
deterministically, from a theory object and a plain vector of candidate DOIs,
regardless of where the DOIs came from.

```{r new-evidence-basic}
theory <- tf_theory("demo", "A demonstration theory") |>
  tf_add_construct("c_arousal", "Arousal", "Bodily activation.")
theory$evidence <- list(list(supports = "p1",
                              source_doi = "10.1016/j.brat.2015.10.002"))

candidates <- c("10.1016/j.brat.2015.10.002",       # already cited
                "https://doi.org/10.1037/0033-2909.99.1.20")  # not yet cited

tf_new_evidence_dois(theory, candidates)
```

The comparison is on a normalised form of each DOI (lowercased, with a
`doi.org`/`dx.doi.org` URL prefix stripped), so a plain DOI and a resolvable
URL for the same work are recognised as the same source. The function takes no
network dependency itself: the search is left entirely to whichever tool
supplies the candidate list.

### Combining it with scopusflow

[`scopusflow`](https://pablobernabeu.github.io/scopusflow/) is a companion R
package, by the same author, for querying the Elsevier Scopus Search API. It
is a natural source of candidate DOIs: `scopus_fetch()` retrieves records for
a query, and `scopus_extract_dois()` reduces them to a plain DOI vector, which
is exactly the input `tf_new_evidence_dois()` expects.

`scopusflow` is not a dependency of `theoryforge`, so the snippet below is
illustrative rather than executed by this vignette. Install `scopusflow` and
supply a Scopus API key (see its `scopus_has_key()`) to run it.

```r
library(scopusflow)

query <- scopus_query("panic disorder", "interoception", .op = "AND")
records <- scopus_fetch(query, years = 2015:2026)
candidates <- scopus_extract_dois(records)

tf_new_evidence_dois(theory, candidates)
```

`scopus_diff_dois()` extends this to tracking a search over time: it compares
an earlier and a later retrieval (either `scopus_records` objects or DOI
vectors) and reports which DOIs were added, removed or unchanged. Re-running a
saved query and passing the `"added"` DOIs into `tf_new_evidence_dois()` gives
a routine for revisiting a theory's evidence base as the literature grows.

```r
records_2025 <- scopus_fetch(query, years = 2015:2025)
records_2026 <- scopus_fetch(query, years = 2015:2026)
diff <- scopus_diff_dois(records_2025, records_2026)
added <- diff$doi[diff$status == "added"]

tf_new_evidence_dois(theory, added)
```

`scopusflow` also offers `scopus_compare_topics()`, which tracks the relative
publication share of several comparison terms against a reference term across
a range of years. This suits comparing a theory against its registered
alternatives on their standing in the literature, using the alternatives'
labels as the comparison terms.

```r
scopus_compare_topics(
  reference_query = "panic disorder",
  comparison_terms = c("cognitive appraisal account", "biological account"),
  years = 2015:2026
)
```

### Using scopusflow for a Scopus-based corpus

`tf_litmap()` and `tf_landscape()` read the `keywords` and `references` fields
of each corpus record, and `scopusflow` supplies both. `scopus_corpus()` takes
the records from `scopus_fetch()` and enriches them, through Abstract
Retrieval, into a tibble of `id`, `title`, `year`, `keywords` (one character
vector of author keywords per record) and `references` (one data frame of
cited works per record, with `id`, `doi`, `title` and other fields).

`tf_fetch_corpus()` (OpenAlex) stays the built-in default because OpenAlex is
free and keyless, so the literature layer works with no setup. Scopus needs an
institutional subscription and an API key, so `scopusflow` is an opt-in source
rather than a dependency. The two packages exchange plain data, a DOI vector or
a corpus written to a file, with no coupling in either direction; that keeps
theoryforge dependency-light and usable out of the box, and lets a reader reach
for whichever index they have access to.

The corpus format expects a top-level `{schema_version, id, records}` envelope
and, within each record, `references` as a flat list of id strings, whereas
`scopus_corpus()` returns a tibble whose `references` entries are data frames.
So build the envelope explicitly, reducing each references data frame to one
id string per cited work (the DOI where present, the Scopus id otherwise),
write the result to disk and read it back with `tf_read_corpus()`:

```r
# illustrative: needs scopusflow and a configured Scopus API key
recs <- scopusflow::scopus_fetch(query, years = 2015:2026)
sc   <- scopusflow::scopus_corpus(recs)   # id, title, year, keywords, references

records <- lapply(seq_len(nrow(sc)), function(i) {
  refs <- sc$references[[i]]              # one data frame of cited works
  ref_ids <- ifelse(is.na(refs$doi), refs$id, refs$doi)
  list(id         = sc$id[[i]],
       title      = sc$title[[i]],
       year       = sc$year[[i]],
       keywords   = as.list(sc$keywords[[i]]),
       references = as.list(ref_ids[!is.na(ref_ids)]))
})
corpus <- list(schema_version = "1.0",
               id = "scopus:panic disorder",
               records = records)
jsonlite::write_json(corpus, "corpus.json", auto_unbox = TRUE)

lit <- tf_read_corpus("corpus.json")
tf_litmap(lit)
```
