---
title: "Mapping the literature"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = FALSE, comment = "")
# Console colour carries no meaning on a rendered page. pkgdown turns it on for
# its own build, and the escape sequences then reach the reader as literal text,
# so colour is switched off here for a plain vignette render and a site build
# alike. The fixed width keeps printed output inside the documentation column.
options(cli.num_colors = 1, cli.hyperlink = FALSE, crayon.enabled = FALSE,
        width = 80)
# Both optional packages are needed to turn the DOT source into an inline
# SVG, so the rendering chunks below are skipped when either is absent.
has_diagram <- requireNamespace("DiagrammeR", quietly = TRUE) &&
  requireNamespace("DiagrammeRsvg", quietly = TRUE)
```

A theory is rarely built in isolation. It sits within a field that already has
crowded areas, sparse areas and recurring sets of references. The literature
layer of `theoryforge` gives a deterministic, machine-checkable account of that
field, so the placement of a theory within it can be argued from evidence
rather than from impression. This article reads a small corpus, computes a
bibliometric map, positions a theory against the map, and emits a diagram. The
analysis functions are deterministic and require no optional dependency.

```{r setup}
library(theoryforge)
```

## Reading a corpus

A corpus is a simple object of the form `{schema_version, id, records}`, where
each record carries an `id` and may carry `title`, `year`, `keywords` and
`references`. The package ships a small fixture corpus around panic and anxiety
research. The format is chosen by file extension, so the same reader handles
YAML and JSON.

```{r read-corpus}
corpus_path <- system.file("fixtures/panic-corpus.yaml",
                           package = "theoryforge")
corpus <- tf_read_corpus(corpus_path)

corpus$id
length(corpus$records)   # number of works in the corpus
corpus$records[[1]]       # first record
```

Keywords and references are strings. An unquoted integer, such as a PubMed or
Scopus id, is read as its decimal string, so `12345678` and `"12345678"` name
the same work in R and in Python. `tf_litmap()` and `tf_landscape()` check the
corpus before counting anything, and a message names the record at fault. They
stop on a missing or misspelt `records` key, a record that is not a mapping and
a fraction or nested list among the values. They also stop on an integer of
2^53 or more, since R cannot hold one exactly. YAML reads an unquoted `NO`
(nitric oxide), `yes`, `on` or `off` as a logical, so quote such a keyword.

## Building a literature map

`tf_litmap()` computes three things from the records. It counts keyword
co-occurrence within records, groups co-occurring keywords into thematic
components, and counts reference co-citation. An edge is kept only when its
co-occurrence count reaches `min_link` (default `2`), which suppresses incidental
pairings. Records are read in file order and every result is sorted, so the map
is fully deterministic.

```{r litmap}
lm <- tf_litmap(corpus, min_link = 2)

lm$n_records
unlist(lm$keywords)                 # the vocabulary of the corpus
lm$keyword_cooccurrence[[1]]        # first kept co-occurrence edge
```

The thematic components are the connected groups of co-occurring keywords. The
fixture is constructed so that the threshold-two graph breaks into four
entirely separate themes.

```{r themes}
for (theme in lm$themes) {
  cat(sprintf("%s: %s\n", theme$id,
              paste(unlist(theme$keywords), collapse = ", ")))
}
```

Connected components suit this fixture, built to fall into separate themes. On
a real corpus, a few keywords that most records carry, such as the name of the
field, link nearly every keyword to every other, so a single component holds the
map. Seven OpenAlex corpora of 100 to 600 records each gave a largest theme
holding at least 98.8 per cent of the linked keywords. When one theme holds more
than half of them, `tf_litmap()` warns that the themes, and any landscape built
on them, are not informative. The map itself is returned unchanged. Below, one
keyword shared by two pairs joins them into one theme.

```{r hub-warning}
hub <- list(schema_version = "1.0", id = "hub", records = list(
  list(id = "w1", keywords = list("anxiety", "arousal")),
  list(id = "w2", keywords = list("anxiety", "arousal")),
  list(id = "w3", keywords = list("anxiety", "avoidance")),
  list(id = "w4", keywords = list("anxiety", "avoidance"))
))
hub_map <- tf_litmap(hub)
length(hub_map$themes)
```

Co-citation is computed the same way over the `references` field. Two cited
sources are linked when at least `min_link` records in the corpus cite them
together.

```{r cocitation}
lm$co_citation
```

Co-citation maps of real corpora are far larger than keyword maps, because each
record cites dozens of works and every pair of them counts. 200 OpenAlex records
on panic disorder give 11,211 co-citation edges at the default threshold of two,
most of them with a count of exactly two, against 616 keyword edges.
`min_cocitation` sets the co-citation threshold on its own, leaving `min_link`
to the keywords. It defaults to `min_link`.

```{r cocitation-threshold}
tf_litmap(corpus, min_link = 2, min_cocitation = 3)$co_citation
```

## Positioning a theory against the field

`tf_landscape()` maps a theory's focal constructs and its registered
alternatives onto the themes from `tf_litmap()`, skipping the co-citation count,
which it does not use. A theme is tagged `"under_theorised"` when neither the
focal theory nor any alternative touches it, `"covered"` when exactly one does, and `"crowded"` when two or more do. Matching
is by shared words between the theme keywords and the construct labels, or the
label and key constructs of an alternative, and the result names the words
behind every match. Three kinds of word never match. Field tokens, the words
most of the corpus shares, are those in the keywords of more than
`max_token_share` of the records, half by default. Phenomenon tokens are the
words of the theory's title, which names the phenomenon that the focal theory
and its rivals all explain. A construct word that also appears in the title
does not match either. The third kind is a fixed list of words that name a kind
of account: theory, model, account, hypothesis, framework and approach, with
their plurals. One shared word is enough for a match.

We build a small theory of panic attacks whose constructs speak to arousal and
to appraisal, and register one alternative that also speaks to appraisal. The
title names the phenomenon, so its words are set aside and the construct labels
decide the matches. This leaves the arousal theme covered, the appraisal theme
crowded and the remaining themes untouched.

```{r theory}
theory <- tf_theory("panic-appraisal", "A theory of panic attacks") |>
  tf_add_construct("c_arousal", "Physiological arousal",
                   "Bodily activation in response to a stressor.") |>
  tf_add_construct("c_appraisal", "Threat appraisal",
                   "Catastrophic appraisal of bodily sensations.") |>
  tf_add_alternative("alt_cog", "Cognitive appraisal account",
                     key_constructs = c("appraisal"))

landscape <- tf_landscape(theory, corpus, min_link = 2)
landscape$theory_id
unlist(landscape$phenomenon_tokens)
```

Each theme in the result records which alternatives land on it, whether the
focal theory lands on it, the resulting status and the words behind each match,
in `focal_terms` and in `alternative_terms`.

```{r landscape-themes}
words <- function(x) if (length(x) > 0L) paste(unlist(x), collapse = ", ") else "none"
for (theme in landscape$themes) {
  alts <- vapply(theme$alternative_terms,
                 function(a) sprintf("%s (%s)", a$id, words(a$terms)), character(1))
  cat(sprintf("%s | status: %s | focal: %s | alternatives: %s\n",
              theme$id, theme$status, words(theme$focal_terms), words(alts)))
}
```

Two summaries draw out the practical reading. The under-theorised fronts are
themes that none of the registered accounts addresses, which says nothing of
accounts the theory does not register. The crowded themes, listed under the key
`redundancy_risk` that 0.6.0 gave them, are addressed by two or more accounts,
so they call for predictions that discriminate between those accounts. They do
not show that any account is redundant.

```{r landscape-summary}
unlist(landscape$under_theorised_fronts)
unlist(landscape$redundancy_risk)
```

## Emitting a literature diagram

`tf_lit_diagram()` renders a deterministic Graphviz DOT string for the
literature layer. The keyword co-occurrence and co-citation maps render as
undirected graphs, and a landscape result renders as a directed theme map. The
output is a plain string, so it can be written to a `.dot` file or passed to a
Graphviz renderer.

```{r diagram-keywords}
cat(tf_lit_diagram(lm, type = "keyword_cooccurrence"))
```

`tf_render_diagram()` accepts a DOT string as well as a theory, so the
literature diagrams render the same way as the theory views.

```{r diagram-keywords-render, results = "asis", eval = has_diagram}
cat('<div class="tf-figure tf-diagram">',
    tf_render_diagram(
      tf_lit_diagram(lm, type = "keyword_cooccurrence"),
      as = "svg"
    ),
    '</div>',
    sep = "")
```

The co-citation map links the references that the corpus cites together, so
tightly co-cited work stands out as the intellectual base the field shares.

```{r diagram-cocitation}
cat(tf_lit_diagram(lm, type = "co_citation"))
```

```{r diagram-cocitation-render, results = "asis", eval = has_diagram}
cat('<div class="tf-figure tf-diagram">',
    tf_render_diagram(
      tf_lit_diagram(lm, type = "co_citation"),
      as = "svg"
    ),
    '</div>',
    sep = "")
```

On a real corpus that diagram can run to thousands of edges and a megabyte of
DOT, which Graphviz cannot lay out legibly. `max_edges` keeps the strongest
edges, the highest counts first with ties going to the pair that sorts first,
and draws only their endpoints. It applies to both network views.

```{r diagram-cocitation-capped}
cat(tf_lit_diagram(lm, type = "co_citation", max_edges = 1))
```

The theme landscape diagram is produced from the `tf_landscape()` result rather
than from the map, since it carries the focal theory and the alternatives.

```{r diagram-landscape}
cat(tf_lit_diagram(landscape, type = "theme_landscape"))
```

```{r diagram-landscape-render, results = "asis", eval = has_diagram}
cat('<div class="tf-figure tf-diagram">',
    tf_render_diagram(
      tf_lit_diagram(landscape, type = "theme_landscape"),
      as = "svg"
    ),
    '</div>',
    sep = "")
```

## Fetching a corpus from OpenAlex

The corpus above was read from a file. A corpus can also be assembled from the
OpenAlex works API with `tf_fetch_corpus()`. This adapter is assistive. It
makes a live network call and its result depends on an external service that
changes over time, so it sits outside the deterministic core of the package.
The call is shown here but not run.

```{r fetch, eval = FALSE}
fetched <- tf_fetch_corpus("panic disorder interoception",
                           per_page = 100,
                           max_records = 400)
fetched$source    # the query, when it ran, how many works matched and were kept
jsonlite::write_json(fetched, "panic-corpus.json", auto_unbox = TRUE,
                     null = "null")
fetched <- tf_read_corpus("panic-corpus.json")
fetched_map <- tf_litmap(fetched)
```

OpenAlex answers requests without a key within a budget of USD 0.10 a day. A
search page costs USD 0.001, so the budget covers about 100 pages. A free API
key raises the budget to USD 1 a day. Store the key in the `OPENALEX_API_KEY`
environment variable, for example in `.Renviron`, and `tf_fetch_corpus()`
sends it in a request header, never in the URL or the corpus. The `mailto`
argument is still accepted, but OpenAlex now ignores it. When OpenAlex refuses
a request, typically with HTTP 429 once the budget is spent, the function
stops with the status and OpenAlex's message. It used to return an empty
corpus, which read as a literature with no themes.

Each request returns one page of works, ranked by relevance, from a result set
that is usually far larger: a search for panic disorder matched about 187,000
works in October 2026. OpenAlex supports pages of up to 100 works and has
deprecated larger ones, although it still accepts 200. `max_records` pages on
through OpenAlex's cursor until that many works are collected, and
`fetched$source` records the query, the UTC time of retrieval, the number of
works that matched and the number kept. That record matters because the
keywords change. Since late September 2026, OpenAlex has written each work's
keywords with a language model that reads its title, abstract and venue. It
merges and splits that vocabulary over time, so the same works fetched a week
apart can carry different keywords. Works left without keywords fall back to
their concepts, a deprecated vocabulary whose capitalised names do not match
the lower-case keywords. Each record also keeps the work's DOI as OpenAlex
gives it. The example writes the corpus with `null = "null"`, because
`jsonlite` would otherwise write a missing DOI or count as an empty object,
which the corpus schema does not allow.

Once fetched, the corpus object has the same shape as a file-read corpus, so it
flows into `tf_litmap()`, `tf_landscape()` and `tf_lit_diagram()` without any
further change. The file-read `corpus` used throughout this article shows that
shape.

```{r fetch-shape}
str(corpus, max.level = 2, list.len = 4)
```

The recommended pattern is to fetch once, write the result to disk, and then
work from the saved file so that later analyses remain reproducible.

## Tracking new evidence with an external search

Locating a corpus is only one use of a literature search. A second, recurring
need is narrower: checking whether a search has turned up any source the
theory does not already cite. `tf_new_evidence_dois()` answers that question
deterministically, from a theory object and a plain vector of candidate DOIs,
regardless of where the DOIs came from.

The bundled panic theory records one DOI as evidence and one for each of its
two registered alternatives, and a DOI cited as an alternative counts as cited
just as an evidence DOI does. Those are the sources the candidate list below is
checked against.

```{r new-evidence-basic}
panic <- tf_read(system.file("fixtures", "panic-network.theory.yaml",
                             package = "theoryforge"))

candidates <- c(
  "10.1016/j.brat.2015.10.002",                   # already cited as evidence
  # already cited as an alternative, in URL form
  "https://doi.org/10.1016/0005-7967(86)90011-2",
  "https://doi.org/10.1037/0033-2909.99.1.20"     # not yet cited
)

tf_new_evidence_dois(panic, candidates)
```

The comparison is on a normalised form of each DOI, the DOI itself found
wherever it sits in the text, lowercased and percent-decoded. A plain DOI, a
resolvable URL and the forms `doi: 10...` and `DOI 10...` for the same work are
therefore recognised as the same source, and a trailing full stop is ignored. The function takes no
network dependency itself: the search is left entirely to whichever tool
supplies the candidate list.

### Combining it with scopusflow

[`scopusflow`](https://pablobernabeu.github.io/scopusflow/) is a companion R
package, by the same author, for querying the Elsevier Scopus Search API. It
is a natural source of candidate DOIs: `scopus_fetch()` retrieves records for
a query, and `scopus_extract_dois()` reduces them to a plain DOI vector, which
is exactly the input `tf_new_evidence_dois()` expects.

`scopusflow` is not a dependency of `theoryforge`, so the snippet below is
illustrative rather than executed by this vignette. Install `scopusflow` and
supply a Scopus API key (see its `scopus_has_key()`) to run it.

```r
library(scopusflow)

query <- scopus_query("panic disorder", "interoception", .op = "AND")
records <- scopus_fetch(query, years = 2015:2026)
candidates <- scopus_extract_dois(records)

tf_new_evidence_dois(panic, candidates)
```

`scopus_diff_dois()` extends this to tracking a search over time: it compares
an earlier and a later retrieval (either `scopus_records` objects or DOI
vectors) and reports which DOIs were added, removed or unchanged. Re-running a
saved query and passing the `"added"` DOIs into `tf_new_evidence_dois()` gives
a routine for revisiting a theory's evidence base as the literature grows.

```r
records_2025 <- scopus_fetch(query, years = 2015:2025)
records_2026 <- scopus_fetch(query, years = 2015:2026)
diff <- scopus_diff_dois(records_2025, records_2026)
added <- diff$doi[diff$status == "added"]

tf_new_evidence_dois(panic, added)
```

The hand-off itself needs nothing from `scopusflow`: `scopus_diff_dois()`
returns one row per DOI with its status, which we reproduce here so the last
step runs. Only the newly added DOI comes back as new evidence.

```{r diff-standin}
diff <- data.frame(
  doi = c("10.1016/j.brat.2015.10.002", "10.1037/0033-2909.99.1.20"),
  status = c("unchanged", "added")
)
added <- diff$doi[diff$status == "added"]

tf_new_evidence_dois(panic, added)
```

`scopusflow` also offers `scopus_compare_topics()`, which tracks the relative
publication share of several comparison terms against a reference term across
a range of years. This suits comparing a theory against its registered
alternatives on their standing in the literature, using the alternatives'
labels as the comparison terms.

```r
scopus_compare_topics(
  reference_query = "panic disorder",
  comparison_terms = c("cognitive appraisal account", "biological account"),
  years = 2015:2026
)
```

The comparison and its plot, with per-year stability bands, are shown in
scopusflow's [Comparing topics](https://pablobernabeu.github.io/scopusflow/articles/comparing-topics.html)
article.

### Using scopusflow for a Scopus-based corpus

`tf_litmap()` and `tf_landscape()` read the `keywords` and `references` fields
of each corpus record, and `scopusflow` supplies both. `scopus_corpus()` takes
the records from `scopus_fetch()` and enriches them, through Abstract
Retrieval, into a tibble of `id`, `title`, `year`, `keywords` (one character
vector of author keywords per record) and `references` (one data frame of
cited works per record, with `id`, `doi`, `title` and other fields).

`tf_fetch_corpus()` (OpenAlex) stays the built-in default because OpenAlex is
free and answers modest use without a key, so the literature layer works with
no setup. Scopus needs an institutional subscription and an API key, so
`scopusflow` is an opt-in source rather than a dependency. The two packages
exchange plain data, a DOI vector or a corpus written to a file, with no
coupling in either direction. That keeps theoryforge dependency-light and
usable straight after install, and lets a reader reach for whichever index
they have access to.

The corpus format expects a top-level `{schema_version, id, records}` mapping
and, within each record, `references` as a flat list of id strings, whereas
`scopus_corpus()` returns a tibble whose `references` entries are data frames.
Fetch the records and enrich them first:

```r
# illustrative: needs scopusflow and a configured Scopus API key
recs <- scopusflow::scopus_fetch(query, years = 2015:2026)
# id, title, year, keywords, references
sc   <- scopusflow::scopus_corpus(recs)
```

`tf_litmap()` compares references and keywords as exact strings, so the
conversion has to give each cited work one spelling across the whole corpus.
Scopus gives a cited work's DOI in one record, in different capitals in
another and not at all in a third, so a key built from the DOI splits one work
into several nodes and the co-citation map comes out empty. Author keywords
keep their authors' capitals, which splits the keyword themes in the same way.
The adapter below therefore keys each cited work by its Scopus identifier, as
`scopus:<id>`. A reference that carries only a DOI borrows the identifier of
another reference with the same DOI anywhere in the corpus, and is keyed as
`doi:<doi>`, with the DOI lower-cased and any resolver prefix removed, when
no such reference exists. References with neither are dropped. Keywords are
trimmed and lower-cased, and a missing year is written as `null`.

```{r scopus-adapter}
scopus_corpus_to_tf <- function(sc, id) {
  # DOIs are case-insensitive over ASCII letters only (DOI Handbook, 2.4), so
  # the fold leaves every other character alone.
  fold_doi <- function(doi) {
    doi <- chartr(paste(LETTERS, collapse = ""), paste(letters, collapse = ""),
                  trimws(doi))
    sub("^(https?://(dx\\.)?doi\\.org/|doi:)", "", doi)
  }
  present <- function(x) {
    x <- trimws(as.character(x))
    x[!is.na(x) & !nzchar(x)] <- NA_character_
    x
  }
  # One (Scopus id, DOI) pair per cited work. A record whose lookup failed or
  # found no references has an empty entry.
  cited <- lapply(sc$references, function(refs) {
    if (!is.data.frame(refs) || nrow(refs) == 0L) {
      return(data.frame(sid = character(), doi = character()))
    }
    data.frame(sid = present(refs$id), doi = fold_doi(present(refs$doi)))
  })
  # Every reference that carries both says which Scopus id a DOI belongs to.
  pairs <- do.call(rbind, c(list(data.frame(sid = character(),
                                            doi = character())), cited))
  pairs <- pairs[!is.na(pairs$sid) & !is.na(pairs$doi), , drop = FALSE]
  pairs <- pairs[!duplicated(pairs$doi), , drop = FALSE]
  sid_of_doi <- stats::setNames(pairs$sid, pairs$doi)

  records <- lapply(seq_len(nrow(sc)), function(i) {
    refs <- cited[[i]]
    sid <- refs$sid
    borrow <- is.na(sid) & refs$doi %in% names(sid_of_doi)
    sid[borrow] <- sid_of_doi[refs$doi[borrow]]
    keys <- ifelse(!is.na(sid), paste0("scopus:", sid),
                   ifelse(!is.na(refs$doi), paste0("doi:", refs$doi), NA))
    keywords <- tolower(present(sc$keywords[[i]]))
    record <- list(id = as.character(sc$id[[i]]),
                   title = present(sc$title[[i]]),
                   year = as.integer(sc$year[[i]]),
                   keywords = I(unique(keywords[!is.na(keywords)])),
                   references = I(unique(keys[!is.na(keys)])))
    # The schema accepts a null year, but a title is a string.
    if (is.na(record$title)) record$title <- NULL
    record
  })
  list(schema_version = "1.0", id = id, records = records)
}
```

Write the result with `null = "null"` and `na = "null"`, so that a missing year
reaches the file as `null` (by default `jsonlite` writes the string `"NA"`), then
read it back with `tf_read_corpus()`. `I()` in the adapter keeps a one-element keyword or
reference list an array under `auto_unbox = TRUE`.

```r
# illustrative: continues from the fetch above
lit <- scopus_corpus_to_tf(sc, "scopus:panic disorder")
jsonlite::write_json(lit, "corpus.json", auto_unbox = TRUE,
                     null = "null", na = "null")

corpus <- tf_read_corpus("corpus.json")
tf_litmap(corpus)
```

The next release of `scopusflow` is planned to write this file itself, with the
same keys, so that the adapter is then needed only for earlier versions.
