---
title: "Mapping the literature"
output: rmarkdown::html_vignette
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

A theory is rarely built in isolation. It sits within a field that already has
crowded areas, sparse areas, and recurring sets of references. The literature
layer of `theoryforge` gives a deterministic, machine-checkable account of that
field, so the placement of a theory within it can be argued from evidence
rather than from impression. This article reads a small corpus, computes a
bibliometric map, positions a theory against the map, and emits a diagram. The
analysis functions are parity-tested against the Python twin and require no
optional dependency.

```{r setup}
library(theoryforge)
```

## Reading a corpus

A corpus is a simple object of the form `{schema_version, id, records}`, where
each record carries an `id` and may carry `title`, `year`, `keywords`, and
`references`. The package ships a small fixture corpus around panic and anxiety
research. The format is chosen by file extension, so the same reader handles
YAML and JSON.

```{r read-corpus}
corpus_path <- system.file("fixtures/panic-corpus.yaml", package = "theoryforge")
corpus <- tf_read_corpus(corpus_path)

corpus$id
length(corpus$records)   # number of works in the corpus
corpus$records[[1]]       # first record
```

## Building a literature map

`tf_litmap()` computes three things from the records. It counts keyword
co-occurrence within records, groups co-occurring keywords into thematic
components, and counts reference co-citation. An edge is kept only when its
co-occurrence count reaches `min_link` (default `2`), which suppresses incidental
pairings. Records are read in file order and every result is sorted, so the map
is fully deterministic.

```{r litmap}
lm <- tf_litmap(corpus, min_link = 2)

lm$n_records
unlist(lm$keywords)                 # the vocabulary of the corpus
lm$keyword_cooccurrence[[1]]        # first kept co-occurrence edge
```

The thematic components are the connected groups of co-occurring keywords. The
fixture is constructed so that the threshold-two graph separates cleanly into
four themes.

```{r themes}
for (theme in lm$themes) {
  cat(theme$id, ":", paste(unlist(theme$keywords), collapse = ", "), "\n")
}
```

Co-citation is computed the same way over the `references` field. Two works are
linked when they cite the same pair of sources at least `min_link` times.

```{r cocitation}
lm$co_citation
```

## Positioning a theory against the field

`tf_landscape()` maps a theory's focal constructs and its registered
alternatives onto the themes from `tf_litmap()`. A theme is tagged
`"under_theorised"` when neither the focal theory nor any alternative touches it,
`"covered"` when exactly one does, and `"crowded"` when two or more do. Matching
is by shared tokens between the theme keywords and the text of the construct
labels, the theory title, and the alternatives.

We build a small theory whose constructs speak to arousal and to appraisal, and
register one alternative that also speaks to appraisal. This leaves the arousal
theme covered, the appraisal theme crowded, and the remaining themes untouched.

```{r theory}
theory <- tf_theory("panic-appraisal", "Arousal and appraisal in panic") |>
  tf_add_construct("c_arousal", "Physiological arousal",
                   "Bodily activation in response to a stressor.") |>
  tf_add_construct("c_appraisal", "Threat appraisal",
                   "Catastrophic appraisal of bodily sensations.") |>
  tf_add_alternative("alt_cog", "Cognitive appraisal account",
                     key_constructs = c("appraisal"))

landscape <- tf_landscape(theory, corpus, min_link = 2)
landscape$theory_id
```

Each theme in the result records which alternatives land on it, whether the
focal theory lands on it, and the resulting status.

```{r landscape-themes}
for (theme in landscape$themes) {
  cat(theme$id,
      "| status:", theme$status,
      "| focal:", theme$focal,
      "| alternatives:", paste(unlist(theme$alternatives), collapse = ", "),
      "\n")
}
```

Two summaries draw out the practical reading. The under-theorised fronts are
themes that the field around this theory has not yet engaged, and so are
candidates for new work. The redundancy risk lists crowded themes, where a new
contribution would have to distinguish itself from existing accounts.

```{r landscape-summary}
unlist(landscape$under_theorised_fronts)
unlist(landscape$redundancy_risk)
```

## Emitting a literature diagram

`tf_lit_diagram()` renders a byte-identical Graphviz DOT string for the
literature layer. The keyword co-occurrence and co-citation maps render as
undirected graphs, and a landscape result renders as a directed theme map. The
output is a plain string, so it can be written to a `.dot` file or passed to a
Graphviz renderer.

```{r diagram-keywords}
cat(tf_lit_diagram(lm, type = "keyword_cooccurrence"))
```

The theme landscape diagram is produced from the `tf_landscape()` result rather
than from the map, since it carries the focal theory and the alternatives.

```{r diagram-landscape}
cat(tf_lit_diagram(landscape, type = "theme_landscape"))
```

These strings are compared byte-for-byte against the Python reference outputs in
the package tests, which guarantees that an R analysis and a Python analysis of
the same corpus agree exactly.

## Building a corpus from OpenAlex

The corpus above was read from a file. A corpus can also be assembled from the
OpenAlex works API with `tf_fetch_corpus()`. This adapter is assistive and
parity-exempt. It makes a live network call, its result depends on an external
service that changes over time, and it is therefore excluded from the
deterministic core, the parity tests, and continuous integration. The call is
shown here but not run, and the supplied email enters the OpenAlex polite pool.

```{r fetch, eval = FALSE}
fetched <- tf_fetch_corpus("panic disorder interoception",
                           per_page = 25,
                           mailto = "me@example.org")
fetched_map <- tf_litmap(fetched)
```

Once fetched, the corpus object has the same shape as a file-read corpus, so it
flows into `tf_litmap()`, `tf_landscape()`, and `tf_lit_diagram()` without any
further change. The recommended pattern is to fetch once, write the result to
disk, and then work from the saved file so that later analyses remain
reproducible.
